Subsidiary Recursion in Coq

### 2022-02-07 paper/paper.tex
final edits

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
Code is in Figure~\ref{fig:wordsBy}.  Recall that
\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination, but
will be permitted by subsidiary recursion.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We emphasize that this is
a derivation within the type theory of Coq, and requires no axioms or
other modifications to Coq, except the \verb|-impredicative-set|
flag. Using this derivation, we present several example functions like
\verb|wordsBy|, and prove theorems about them.  A nice example is a
definition of run-length encoding using \verb|span| as a subsidiary
recursion, where we prove that encoding and then decoding returns the
original list.  Our approach applies to the standard datatypes in the
Coq library, and does not require switching libraries or datatype
definitions.

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  With
this approach, one describes transformations to be performed on data
as \emph{algebras}, which can then be \textit{folded} over data.  The
simplest form of algebras, namely $F$-algebras for functor $F$ (called
the \emph{signature functor} of the datatype), are morphisms from
$F\ A$ to $A$, for carrier object $A$.  From a programming
perspective, an $F$-algebra is given input of type $F\ A$, and must
compute a result of type $A$.  An example of $F$ is the signature
functor for lists, which we will use below:

@@

We have previously dubbed this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using certain operations.  Other examples are the
ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
traveling at the speed of light.  Staying in the recursion universe is
good, because we may recurse (via \verb|rec|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type
\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
these details next.

@@

In this paper, we use just the specialization to the case of lists,
with signature functor \verb|ListF A| shown above.  The parameter
\verb|A| is the type for the list elements.  \verb|List| is then
defined to be \verb|Subrec|, with \verb|F| instantiated to
\verb|ListF A|. In general, to use our development to get subsidiary
recursion over some datatype, one must define a signature functor for
the datatype.  Note that \verb|List| is different from the type
\verb|list| of lists in Coq's standard library.  Our development is
meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (definitions elided).

\subsection{Algebras for subsidiary recursion}

\verb|Subrec.v| also implements the notion of algebra we introduced
informally above.  The central definitions are in
Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
type-constructor for algebras, as we see in the definition of
\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|, which
implements our retractive-positive recursive types
(Section~\ref{sec:mu}) at kind \verb|KAlg|.

We need a fixed-point here because \verb|Alg| occurs in the definition
of \verb|Algf|.  This is an essential circularity, because we are
trying to express that algebras take in \verb|fold| functions, which
themselves may accept algebras.  The variable \verb|Alg| occurs
negatively in \verb|FoldT Alg R| which occurs negatively in
\verb|AlgF Alg X|.  Hence it occurs positively in \verb|Algf|, though not strictly
positively.  So we can indeed take a fixed-point of \verb|AlgF| to
define the constant \verb|Alg|.

@@

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the following
operations.  We have a local
\texttt{fold} function, which will allow us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  Then there is \verb|rec|, for recursive calls on data of
type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results as having the type \verb|C| outside
the recursion.  The function \verb|fold| (Figure~\ref{fig:algf}) initiates
top-level folds.  We also can have functions 
between \verb|Alg| and its \verb|Algf|-unfolding.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

Finally, for a recursion scheme, one would like to see not just the
typed interface, but also the computation law.  This is shown as a
theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
to invoking the algebra on \verb|fold|
for the fold function; an invocation of \verb|fold| with the algebra
for the \verb|rec| function; and \verb|d| for the subdata structure.

@@

This first example does not invoke subsidiary recursions, but will
itself be used as a subsidiary recursion in other examples to follow.
Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given a type \verb|R : Set| along with a
fold function for that type (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we will map an input list of
type \verb|R| to a pair of type \verb|list A * R|.  The first
component of this pair is going to be built up from scratch, and so
cannot have type \verb|R|; we cannot statically ensure that outer
recursions on it are legal.  But the second component will always be a
subdatum of the input list, and so can still have type \verb|R|,
enabling outer recursive calls.  So we want:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will return \verb|SpanNoMatch| to signal that it is in the
one tricky case where \verb|p| does not match the first element.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} shows the algebra
\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
\verb|SpanF| of the required type \verb|Set -> Set|.  We use
\verb|rollAlg| to create an algebra from something whose type is an
application of \verb|AlgF|.  This takes in all the components of the
recursion universe: the abstract type \verb|R|, the fold function
(\verb|fo|) for any subsidiary recursions (not needed here), a
function we choose to name \verb|span| for making recursive calls, and
finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
\verb|xs|.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  For either possible return
value of \verb|span tl|, we add the head to the front of the
prefix.

@@

\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
Figure~\ref{fig:span}.  This function invokes the fold function it is
given, on \verb|SpanAlg|.  The final twist is now in the definition of
\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
return \verb|([],xs)|, which we can do here, because we have
\verb|xs : R|.  It was only inside the algebra that we lost the
information that the subdata structure of type \verb|F R| is derived
from a value of type \verb|R|.  If \verb|spanhr| returns
\verb|SpanSomeMatch l r|, then we return the nonempty
prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| 
for subsidiary recursion (e.g., in \verb|wordsBy|, below).

@@

The code for \verb|WordsByAlg| is essentially
the same as what we saw in Section~\ref{sec:intro}.  Recall that this function drops elements
that satisfy \verb|p|, and returns the list of
sublists between maximal sequences of such elements.  The algebra
pattern-matches on \verb|xs : ListF A R|. In the
\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we
drop it and recurse.  Legality of the recursive call follows by typing: \verb|tl : R|
has the type expected by \verb|wordsBy : R -> list (list A)|.  Otherwise,
we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
does not satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
legal to apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.  The figure also shows the code for the subsidiary
recursion \verb|wordsByr|.

\begin{figure}
\begin{minted}{coq}
  Definition WordsByAlg(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
\end{minted}
\caption{The algebra \texttt{WordsByAlg}, and functions \texttt{wordsBy} and \texttt{wordsByr} folding it (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

This example shows how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra|
defines a function \verb|repeatedly| in essentially the same way as
\verb|mapThrough| in Figure~\ref{fig:mapthroughhs} (we propose a more
informative name).  This function behaves like the standard \verb|map|
function on lists, except that the function \verb|f| that we are
mapping (or ``mapping through'') takes in not just the current element
\verb|a|, but also the tail \verb|as|.  It then returns the value
\verb|b| to include in the output list, and whatever other list it
wishes, upon which \verb|mapThrough| will then recurse.

@@

Given this definition, the Coq definition of \verb|mapThrough| is shown in Figure~\ref{fig:mapthrough}.  
\verb|MapThroughAlg| is similar to the
Haskell code above, though when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  
From the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|, so we may indeed invoke \verb|mapThrough : R -> list B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

@@

As we have seen, our definitions require a form of positive-recursive
types, to allow algebras to accept fold functions that themselves
require algebras, and also for the definition of \verb|Subrec| (which
we will see in more detail in the next section).  Full
positive-recursive types are incompatible with Coq's type
theory~\cite{coquand88}.  One can impose some restrictions on large
eliminations which then enable positive-recursive
types~\cite{blanqui05}, but this requires changing the underlying
theory.  Here we exploit Coq's
impredicative polymorphism to obtain a different solution.

Our starting point is a type scheme \verb|F : Set -> Set|, with an \verb|fmap| function
(morphism part of the functor) of type
\begin{minted}{coq}
forall A B : Set, (A -> B) -> F A -> F B
\end{minted}
\noindent which satisfies the identity-preservation law for functors:
\begin{minted}{coq}
fmapId : forall (A : Set)(d : F A), fmap (fun x => x) d = d
\end{minted}
\noindent Then we make the definitions of Figure~\ref{fig:mu}.  The critical
idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
to have a definition like
\begin{minted}{coq}
  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
\end{minted}
\noindent This is exactly what is used in many approaches to modular
datatypes in functional programming, like
Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
rejected by Coq, as instantiations of
\verb|F| that are not strictly positive would be unsound.

@@

For a simple example: suppose we instantiate \verb|F| with
\verb|ListF A| (from Section~\ref{sec:recu}).  Our derivation uses a
different type that wraps \verb|F|, but using \verb|ListF A|
demonstrates the issue in a simple form.  Let us temporarily define
\verb|List A| as \verb|Mu (ListF A)| (again, for subsidiary recursion
do not use just \verb|ListF| directly).  The canonical way to define
the empty list would be:
\begin{minted}{coq}
Definition mkNil := mu (List A) (fun x => x) (NilF A)
\end{minted}
\noindent But given this, there are infinitely many other equivalent
definitions.  For any \verb|Q : Set|, we could take
\begin{minted}{coq}
Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
\end{minted}
\noindent Since \verb|fmap f (NilF A)| equals \verb|NilF B| for
\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
definitionally nor provably.  Of course, one could define a function
that puts \verb|Mu| values in canonical form by folding \verb|inMu|
over them.  Then \verb|mkNil| and \verb|mkNil'| would be equivalent.
But they would still not be provably equal, which is the problem of
noncanonicity.  We will see how to work around this in
Section~\ref{sec:examplesi}.  First, though, let us complete the
exposition of our implementation of subsidiary recursion.

@@

The rest of the interface for \verb|Subrec| is shown in
Figure~\ref{fig:subrecb}.  To \verb|fold| an algebra
\verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
\verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| the
definition of \verb|Subrec| and apply that to the algebra (with its
carrier).

>>>>

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
Code is in Figure~\ref{fig:wordsBy}.  Recall that
\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination, but
is permitted by subsidiary recursion.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We emphasize that this is
a derivation within the type theory of Coq, and requires no axioms or
other modifications to Coq, except the \verb|-impredicative-set|
flag. Using this derivation, we present several example functions like
\verb|wordsBy|, and prove theorems about them.  A nice example is a
definition of run-length encoding using \verb|span| as a subsidiary
recursion, where we prove that encoding and then decoding returns the
original list.  Our approach does not require switching libraries or
datatype definitions.

@@

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  With
this approach, one describes transformations to be performed on data
as \emph{algebras}, which can then be \textit{folded} over data.  The
simplest form of algebras, namely $F$-algebras for functor $F$ (called
the \emph{signature functor} of the datatype), are morphisms from
$F\ A$ to $A$, for carrier object $A$.  From a programming
perspective, an $F$-algebra is given input of type $F\ A$, and must
compute a result of type $A$.  An example of $F$ is the signature
functor for lists, paramatrized by the type \verb|A| of elements:

@@

We have previously dubbed interfaces for recursion \emph{recursion
universes}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|), which one
cannot escape using certain operations.  Other examples are the
ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
traveling at the speed of light.  Staying in the recursion universe is
good, because we may recurse (via \verb|rec|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type
\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
these details next.

@@

In this paper, our examples use \verb|ListF A| (shown above) to
instantiate the parameter \verb|F|.  \verb|List| is then defined to be
\verb|Subrec|. In general, to use our development to get subsidiary
recursion over some datatype, one must define a signature functor for
the datatype.  Note that \verb|List| is different from the type
\verb|list| of lists in Coq's standard library.  Our development is
meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (definitions elided).

\subsection{Algebras for subsidiary recursion (\texttt{Subrec.v})}

Let us now look more formally at the notion of algebra we introduced
informally above.  The central definitions are in Figure~\ref{fig:algf}.
\verb|KAlg| is the kind for the type-constructor for algebras, as we
see in the definition of \verb|Alg|.  This type-constructor \verb|Alg|
is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken
using \verb|MuAlg|, which implements our retractive-positive recursive
types (Section~\ref{sec:mu}) at kind \verb|KAlg|.

We need a fixed-point here since \verb|Alg| occurs in the definition
of \verb|Algf|.  This is an essential circularity, because we are
trying to express that algebras take in \verb|fold| functions, which
themselves may accept algebras.  The variable \verb|Alg| occurs
negatively in \verb|FoldT Alg R| which occurs negatively in
\verb|AlgF Alg X|.  Hence it occurs positively in \verb|Algf|, though not strictly
positively.  So we can indeed take a fixed-point of \verb|AlgF| to
define the constant \verb|Alg|.

@@

Let us look further at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the 
operations that follow.  We have a local
\texttt{fold} function, which will allow us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  Then there is \verb|rec|, for recursive calls on data of
type \verb|R|.  Finally, we have the subdata structure \verb|d : F R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  When we fold an
algebra using a fold function (either global or local) of type
\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
and (ii) outside that recursion, the result will have type \verb|X C|.
Having a functor for the carrier of the algebra gives us the
flexibility to type results inside a recursion with the abstract type
\verb|R|, but view those results with type \verb|C| outside
the recursion.  The function \verb|fold| (Figure~\ref{fig:algf}) initiates
top-level folds.  We also have functions \verb|rollAlg| and \verb|unrollAlg|
between \verb|Alg| and its \verb|Algf|-unfolding.  We will
fill in missing definitions (for \verb|Subrec|, \verb|inn|, \verb|fold|, etc.) in Section~\ref{sec:deriv}.

Finally, for a recursion scheme, one would like to see not just the
typed interface, but also the computation law.  This is shown as a
theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
to invoking the algebra on \verb|fold|
for the fold function; an invocation of \verb|fold| with the algebra
for the \verb|rec| function; and \verb|d| for the subdata structure.
(\verb|FmapId| expresses the identity law for functors, needed
for the proof.)

@@

Our first example is \verb|span| discussed above.  It does not invoke
subsidiary recursions, but will be used as a subsidiary
recursion in later examples.  Given a predicate
\verb|p : A -> bool|, and a value of type \verb|List A|, we would like
to compute a pair of type \verb|list A * List A|, where the first
component is the maximal prefix whose elements satisfy \verb|p|, and
the second is the remaining suffix.  This is the typing for a
top-level recursion.  More generally, though, given a type
\verb|R : Set| along with a fold function for that type (i.e., of type
\verb|FoldT (Alg (ListF A)) R|), we will map an input list of type
\verb|R| to a pair of type \verb|list A * R|.  The first component of
this pair is going to be built up from scratch, and so cannot have
type \verb|R|; we cannot statically ensure that outer recursions on it
would be legal.  But the second component will always be a subdatum of the
input list, and so can still have type \verb|R|.  The typing is:
\begin{minted}{coq}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{minted}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{minted}{coq}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{minted}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of the following type \verb|SpanF R| (with \verb|X| implicit for the constructors):
\begin{minted}{coq}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{minted}
\noindent The idea is that the recursion will return \verb|SpanNoMatch| to signal that it is in the
one tricky case where \verb|p| does not match the first element.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

Figure~\ref{fig:spanalg} shows the algebra
\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  We are defining
an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
\verb|SpanF| of the required type \verb|Set -> Set|.  We use
\verb|rollAlg| to create an algebra from something whose type is an
application of \verb|AlgF|.  This takes in all the components of the
recursion universe: the abstract type \verb|R|, the fold function
\verb|fo| for any subsidiary recursions (not needed here), a
function we choose to name \verb|span| for making recursive calls, and
finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
\verb|xs|.  In the cases where it is empty or where its head
(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
This signals to the caller that we really wished to return
\verb|([],xs)|, but could not because we do not have \verb|xs| at type
\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
tail (\verb|tl : R|) by calling the provided
\verb|span : R -> SpanF R|.  If \verb|span tl| returns
\verb|SpanNoMatch|, that means that we should make \verb|tl| the
suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
have \verb|tl : R| here, so we can do this.  For either possible return
value of \verb|span tl|, we add the head to the front of the
prefix.

@@

\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
Figure~\ref{fig:span}.  This function invokes the fold function it is
given, on \verb|SpanAlg|.  The final twist is now in the definition of
\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
return \verb|([],xs)|, which we can do here, because we have
\verb|xs : R|.  It was only inside the algebra that we lost the
information that the subdata structure of type \verb|F R| is derived
from a value of type \verb|R|.  If \verb|spanhr| returns
\verb|SpanSomeMatch l r|, then we return the nonempty
prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version \verb|breakr| 
for subsidiary recursion.

@@

The code for \verb|WordsByAlg| is essentially the same as
\verb|wordsBy| in Haskell, which we saw in Section~\ref{sec:intro}.
Recall that we must drop elements that satisfy \verb|p|, and return
the list of sublists between maximal sequences of such elements.  The
algebra pattern-matches on \verb|xs : ListF A R|. In the \verb|Cons|
case, if the head (\verb|hd|) satisfies the predicate, then we drop it
and recurse.  Legality of the recursive call follows by typing, since
\verb|tl : R| has the input type required by
\verb|wordsBy : R -> list (list A)|.  Otherwise, we use \verb|breakr|
to obtain the maximal prefix \verb|w| of \verb|tl| that does not
satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
comes from the type \verb|FoldT (ListF A) Alg R| of
\verb|fo|.  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
legal to apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.  The figure also shows the code for the subsidiary
recursion \verb|wordsByr| and top-level recursion \verb|wordsBy|.

\begin{figure}
\begin{minted}{coq}
  Definition WordsByAlg(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
Definition wordsBy(p : A -> bool)(xs : List A) : list (list A) :=
  wordsByr (fold (ListF A)) p xs.
\end{minted}
\caption{The algebra \texttt{WordsByAlg}, and functions \texttt{wordsBy} and \texttt{wordsByr} folding it (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

This example shows how to write a combinator that factors out a
subsidiary recursion.  The Haskell library \verb|Data.List.Extra|
defines a function \verb|repeatedly| in essentially the same way as
\verb|mapThrough| in Figure~\ref{fig:mapthroughhs} (we propose a more
informative name).  This function behaves like the standard \verb|map|
function on lists, except that the function \verb|f| that we are
mapping (or ``mapping through'') takes in not just the current element
\verb|a|, but also the tail \verb|as|.  It then returns the value
\verb|b| to include in the output list, and whatever sublist it
wishes, upon which \verb|mapThrough| will then recurse.

@@

Given this definition, Coq code for \verb|mapThrough| is shown in Figure~\ref{fig:mapthrough}.  
\verb|MapThroughAlg| is similar to the
Haskell code in Figure~\ref{fig:mapthroughhs}, though when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  
From the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|, so we may indeed invoke \verb|mapThrough : R -> list B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

@@

As we have seen, our definitions require a form of positive-recursive
types, to allow algebras to accept fold functions that themselves
require algebras, and also for the definition of \verb|Subrec| (which
we will see in more detail in the next section).  Full
positive-recursive types are incompatible with Coq's type
theory~\cite{coquand88}.  One can impose some restrictions on large
eliminations which then enable positive-recursive
types~\cite{blanqui05}, but this requires changing the underlying
theory.  Here we derive a different solution.

Our starting point is a type scheme \verb|F : Set -> Set|, with an \verb|fmap| function
(morphism part of the functor) of type
\begin{minted}{coq}
forall A B : Set, (A -> B) -> F A -> F B
\end{minted}
\noindent which satisfies the identity-preservation law for functors:
\begin{minted}{coq}
fmapId : forall (A : Set)(d : F A), fmap (fun x => x) d = d
\end{minted}
\noindent Then we make the definitions of Figure~\ref{fig:mu}.  The critical
idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
to use this definition:
\begin{minted}{coq}
  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
\end{minted}
\noindent This is exactly what is used in many approaches to modular
datatypes in functional programming, like
Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
rejected by Coq, as instantiations of
\verb|F| that are not strictly positive would be unsound.

@@

For a simple example of noncanonicity: suppose we instantiate \verb|F| with
\verb|ListF A| (from Section~\ref{sec:recu}).  Our derivation actually uses a
different type that wraps \verb|F|, but using \verb|ListF A|
demonstrates the issue in a simple form.  Let us temporarily define
\verb|List A| as \verb|Mu (ListF A)| (again, for subsidiary recursion
do not use just \verb|ListF| directly).  The canonical way to define
the empty list would be:
\begin{minted}{coq}
Definition mkNil := mu (List A) (fun x => x) (NilF A)
\end{minted}
\noindent But given this, there are infinitely many other equivalent
definitions.  For any \verb|Q : Set|, we could take
\begin{minted}{coq}
Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
\end{minted}
\noindent Since \verb|fmap f (NilF A)| equals \verb|NilF B| for
\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
definitionally nor provably.  Of course, one could define a function
that puts \verb|Mu| values in canonical form by folding \verb|inMu|
over them.  Then \verb|mkNil| and \verb|mkNil'| would be equivalent.
But they would still not be provably equal, which is the problem of
noncanonicity.  We will see how to work around this in
Section~\ref{sec:examplesi}.  First, though, let us complete the
exposition of our implementation of subsidiary recursion.

@@

The rest of the interface for \verb|Subrec| is shown in
Figure~\ref{fig:subrecb}.  To \verb|fold| an algebra
\verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
\verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| \verb|d| and apply that to the algebra (with its
carrier).
